---
title: "Week3 Mon Lab"
author: "Naomi"
date: "2023-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
# ESP 106 Lab 3

Monday, January 23, 2023

## Graphing

In this lab session, we will start by reading merging in data on economic development and indoor and outdoor air pollution. Then we will practice making some graphs with it.

### Question 1

Read in the three csv files: gdppercapitaandgini and airpollution

Both data sets are from Our World in Data: ourworldindata.org

-   The GDP data set has GDP per capita and the GINI index (a measure of income inequality: <https://en.wikipedia.org/wiki/Gini_coefficient>)

-   The air pollution data set has death rates from indoor and outdoor air pollution - units are in deaths per 100,000 people

    -   Indoor air pollution is the **Household Air Pollution from Solid Fuels**

    -   Outdoor air pollution is split into **Particulate Matter** and **Ozone**

gdppercapitaandgini and airpollution csv datasets are imported and renamed :

-   gdppercapitaandgini becomes **gdpgini**
-   airpollution becomes **apol**

```{r}
View(gdpgini)
View(apol)
```

*The column names are long and cumbersome (because they contain information about units, etc.) - you might want to rename some of the columns to make them easier to work with.*

In that case, the columns are renamed to make them simpler.

```{r}
colnames(apol) <- c("Country","Code","Year","PM","Fuels","Ozone","Total")
colnames(gdpgini) <- c("Country","Code","Year","Population","Continent","GINI","GDP")
```

