---
title: "ESP106 Final Project"
author: "Naomi"
date: "2023-03-06"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ESP 106 Final Project

## Question 

Do Metropolitan Statistical Area (MSA) neighborhood status affect
households' transportation mode choice?

## Objective 

The project aims to compare the transportation mode choice between
households living in MSA category defined in NHTS.

## Variables 

### Independent variables: 

Metropolitan Statistical Area (MSA) neighborhood status:

1.  MSA of 1 million or more population, with rail infrastructure

2.  MSA of 1 million or more population, without rail infrastructure

3.  MSA less than 1 million

4.  Not in MSA

### Dependent variables: 

Transportation mode choices:

1.  Car

2.  Bus

3.  Train

4.  Taxi

5.  Bike

6.  Walk

## Data Source

[National Household Travel Survey 2017](https://nhts.ornl.gov/)

[Data Summary of NHTS on
GitHub](https://github.com/Westat-Transportation/summarizeNHTS.git)

## Workflow

Install 'devtools' package to download NHTS data set from GitHub

```{r}
install.packages('devtools')
devtools::install_github('Westat-Transportation/summarizeNHTS')
```

Download the NHTS 2017 data set to personal GitHub cloud storage

```{r}
library(summarizeNHTS)
download_nhts_data("2017", exdir="C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\final")
dataset <- read_data("2017", csv_path="C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\final")
statistic <- summarize_data(
    data = dataset,
    agg = "household_count",
    by = c("HHSIZE","HHVEHCNT")
)
make_chart(statistic)
```
