---
title: "Lab 5"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\W5 Lab\\") 
#use this to set your working directory so knitr understands it
```

## Lab 5

### Due Tuesday Feb1st - Recommended to complete this before starting the midterm

This lab we will look at some data from the plastic trash picked up
during clean-up events around the world. I took this data set from the
Tidy Tuesday website. You can read the documentation
[here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-01-26/readme.md),
including the references and description of the different column names.

I have done some pre-processing of the data for you for this lab, to
create two more easy-to-use data frames.

First read in the countrytotals.csv data frame

```{r}
setwd("C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\W5 Lab\\ESP106_week5_data\\")
tidytue <- read.csv("countrytotals.csv")

```

Have a look at the data frame. Then column "total" gives the total
number of pieces of plastic picked up in that country in 2020. The
columns "num_events" and "volunteers" give the number of trash pick-up
events and the number of volunteers in that country. We are going to use
this to investigate where the plastic trash problem is worst.

1.  What 5 countries had the worst plastic problem as measured by the
    number of pieces of trash picked up?

```{r echo=FALSE}

#hint order(datacolumn, decreasing=TRUE) will give you the ranking of the countries by the values in the data column
#you can use this to re-order the country data frame, ordering it from the country with the most trash picked up to the least
```

**Answer:**

```{r}
head (tidytue[order(tidytue$total, 
                   decreasing=TRUE),])
```

[ Another way to show this using dplyr package is by using top_n
function to pick 5 countries with highest "total" value, then using
arrange function to sort the countries in descending order.
]{style="color:orange"}

```{r}
library(dplyr)

top5 <- arrange(top_n(tidytue, 5, total), desc(total))
top5

```

[ Both ways show that 5 countries that have the worst plastic problem
are: ]{style="color:orange"}

[ 1. Nigeria ]{style="color:orange"}

[ 2. Philippines ]{style="color:orange"}

[ 3. Switzerland ]{style="color:orange"}

[ 4. India ]{style="color:orange"}

[ 5. Togo ]{style="color:orange"}

2.  Make a plot showing the distribution of volunteers across countries

```{r echo=FALSE}
library(ggplot2)

tidytue$country <- gsub('United States of America', 'USA', tidytue$country)
tidytue$country <- gsub('United Kingdom of Great Britain & Northern Ireland', 'UK', tidytue$country)
tidytue$country <- gsub('United Arab Emirates', 'UAE', tidytue$country) 


ggplot(tidytue, aes(x=country, y=volunteers)) + 
    geom_bar(stat = "identity") +
    labs(title="Trash Picking-up Volunteers around the World", y = "Number of Volunteers", x = "Country") +
    coord_flip()

```

3.  Notice that there is a lot of variation across countries in the
    number of volunteers involved in trash pickup. What problem might
    that cause for the interpretation of your answer to question 1?

```{r}
summary(tidytue$volunteers)
```

**Answer:** [ As shown from the summary of Tidy Tuesday's Volunteers
above, it is seen that the mean of 39,330 cannot represent the number of
volunteers involved across countries because the range is to wide with
minimum value of 12 and maximum value of 703,165.
]{style="color:orange"}

4.  Add a column to the data frame creating a variable that should be
    more closely related to the presence of plastic pollution in the
    country

```{r echo=FALSE}
ratio <- tidytue$total/tidytue$num_events
tidytue <- cbind(tidytue, ratio)

```

[To better illustrate the the condition of plastic pollution in each
countries, I add the variable of number of plastic trash picked up per
picking-up events under the column of 'ratio'. ]{style="color:orange"}

5.  What 5 countries have the worst plastic pollution, as measured by
    this new variable?

```{r echo=FALSE}
head (tidytue[order(tidytue$ratio, 
                   decreasing=TRUE),])

```

**Answer:**

[After including the new variable of average number of trash picked up
per events, 5 countries that have the worst plastic pollution are:
]{style="color:orange"}

[ 1. Togo ]{style="color:orange"}

[ 2. Burkina Faso ]{style="color:orange"}

[ 3. Germany ]{style="color:orange"}

[ 4. El Salvador ]{style="color:orange"}

[ 5. Ghana ]{style="color:orange"}

Now we will make a plot of the variation in the types of trash and how
it differs around the world. Read in the continenttypes.csv data frame.
This gives the breakdown of the different types of plastic collected on
each continent in 2020 and the total number of pick up events.

```{r}
setwd("C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\W5 Lab\\ESP106_week5_data\\")
contype <- read.csv('continenttypes.csv')

library(ggplot2)
ggplot(contype, aes(fill=plastic_type, y=total, x=continent)) + 
    geom_bar(position="dodge", stat="identity") +
    labs(title="Type of Plastic Trash around the World", y = "Numbers of Plastic Trash", x = "Continent") +
    scale_fill_discrete(name = "Plastic Type", labels = c("Empty", "HDPE", "LDPE", "Other", "Polyester ", "Polypropylene", "Polystyrene", "PVC "))
```

6.  Add a column to this data frame with a variable that captures the
    existence of different types of plastic trash, controlling for the
    intensity of the pick-up effort in different continent

```{r echo=FALSE}
te <- contype$total/contype$events
contype <- cbind(contype, te)
contype[, 6] <- round(contype[, 6], digits = 2)
```

[ I add new column of 'te' which shows how many plastic trash (for each
type) picked up per event. ]{style="color:orange"}

7.  Make a plot using ggplot showing both the total amount and
    distribution of types of plastic picked up in each continent in the
    average pick-up event.

Hint: Check out options in the [R graph
gallery](https://www.r-graph-gallery.com)

```{r echo=FALSE}
library(ggplot2)
ggplot(contype, aes(fill=plastic_type, y=te, x=continent)) + 
    geom_bar(position="stack", stat="identity") +
    labs(title="Type of Plastic Trash Picked Up around the World", y = "Numbers of Plastic Trash picked up per Event", x = "Continent") +
    scale_fill_discrete(name = "Plastic Type", labels = c("Empty", "HDPE", "LDPE", "Other", "Polyester ", "Polypropylene", "Polystyrene", "PVC "))

```

8.  Try uploading your R markdown file and plot to your Git Hub
    repository. Don't put your knitted HTML file in Github - these are
    large files that are not designed to be stored on Github
