---
title: "Midterm One"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r}
library(knitr)
```

In this midterm we will analyze some data on the conservation status of
species in North America and spending under the Endangered Species Act.

Answer the following questions by using chunks of R code. Comment on
what your code does. Make sure to add informative axis titles and, where
appropriate, units to your answers. Upload the R markdown file and
knitted output to Canvas.

We will use the file `conservationdata.csv`. This data set has
information on North American species. It has five variables that are
described in the table below.

```{r, echo=FALSE}
tab <- cbind(Name=c("speciesid", "speciesname", "taxon", "conservation status", "listed"), Description = c("unique ID", "scientific name", "Species group", "Conservation status in North America, according to   [NatureServe](https://www.natureserve.org/conservation-tools/conservation-rank-calculator): 1 = Critically Imperiled; 2 = Imperiled; 3 = Vulnerable; 4 = Apparently Secure; 5 = Secure; UNK = Unknown; Prob. Extinct = Probably Extinct; Extinct", "Is the species listed as threatened or endangered under the US Endangered Species Act: 0 = No; 1 = Yes"))

knitr::kable(tab, caption ='Table 1. Variables in "consevationdata.csv"' )
```

Read the file `conservationdata.csv`

```{r}
setwd("C:\\Users\\Benny Panjaitan\\Documents\\GitHub\\esp106-Naomi\\Midterm")
consdata <- read.csv("conservationdata.csv")
```

1.  What fraction of species in the data set are listed under the
    Endangered Species Act? (2 points)

```{r}
all <- nrow(consdata)
end <- nrow(consdata[consdata$listed == '1',])

all/end
end/all*100
```

<span style="color:orange">
The fraction of species listed under the Endangered Species Act is
**1/33.18367**, which means there is *1 endangered species for every
33.18367 species*. Or, there are **3.01353%** species are listed under
the Endangered Species Act.
</span>

2.  Show how many (absolute and relative) species there are for each
    taxonomic group by making a data.frame in which the first column has
    the taxonomic groups, the second column is the number of species in
    that group, and the third column is the percentage of all species in
    that group.

```{r}
tx <- as.data.frame(table(consdata$taxon))
tx <- cbind(tx, tx$Freq/all*100)
colnames(tx) <- c("Taxonomic Groups", "Number of Species", "Percentage (%)")
tx
```

3a) One interesting question is how the conservation status varies
between different taxonomic groups. Make a plot showing the relative
distribution of conservation status within each taxonomic group. There
should be descriptive legend (with words, not with the numeric codes) (3
points)

You can use a "base" plotting method, or ggplot. If you are using
ggplot, stat="count" (counts up and plots the number of observations,
i.e. species, within each group) and position="fill" might both be
useful.

```{r}
ts <- data.frame (consdata$taxon, consdata$conservation_status)
colnames(ts) <- c("Taxonomic Groups", "Conservation Status")

library(ggplot2)
ggplot(ts, aes(`Taxonomic Groups`, fill=`Conservation Status`)) + 
    geom_bar(position="fill", stat="count") +
    labs(title="Relative Distribution of Conservation Status for each Taxonomic Group", y = "Relative Numbers") +
    scale_fill_discrete(labels = c("Critically Imperiled", "Imperiled", "Vulnerable", "Apparently Secure", "Secure", "Extinct", "Probably Extinct", "Unknown", "No Data"))

```

3b) Based on this graph, what is something we might be concerned about
in terms of analyzing the data on conservation status, particularly for
fungi and invertebrates? (1 point)

**Answer:** 
<span style="color:orange">
There is a big proportion of Unknown (UNK) conservation status on Fungi and Invertebrates groups.
</span>

Read in the second data file: `spendingdata.csv`

```{r}
spend <- read.csv("spendingdata.csv")
```

This data set has a species ID that matches the species ID in the
conservation data set (speciesid), year, and the spending on conservation
of that species (expressed in 2015 dollars, i.e., accounting for
inflation)

4a) Make a plot showing the distribution of spending in the year 2016 (3
points)

```{r}
sp2016 <- subset(spend, Year == "2016")

ggplot(sp2016, aes(x=spending)) +
  geom_density(fill="darkgreen", color="grey", alpha=0.8) +
  labs(title="Distribution of Spending in 2016", x="Spending in 2016", y="Spending Density")
```

4b) Notice the (very) long right tail on spending data - we spend a lot
on a very small number of species. Show the IDs of the 3 species with
the most spending in 2016. (2 points)

```{r}
head (sp2016[order(sp2016$spending, decreasing=TRUE),])
```

<span style="color:orange">
The 3 Species ID with most spending in 2016 are **#1632**, **#4486**, and **#1684**
</span>

5.  Merge in the data from the conservation status data frame to the
    spending data frame, so that we have information on species names,
    taxonomic group, and conservation status with the spending data. (2
    points); and use that to show the scientific names of the three
    species identified above.

```{r}
cosp <- merge(consdata, spend, by="speciesid")

top1 <- cosp[which(cosp$Year == "2016" & cosp$speciesid == "1632"),]
top2 <- cosp[which(cosp$Year == "2016" & cosp$speciesid == "4486"),]
top3 <- cosp[which(cosp$Year == "2016" & cosp$speciesid == "1684"),]

spc <- data.frame()
spc <- rbind.data.frame(top1, top2, top3)
spc

```

Look up these scientific names - what is the common name for these
species?

**Answer:** 
```{r}
spc <- subset(spc, select=c(1,2))
name <- c("Chinook salmon", "Rainbow trout", "Coho salmon")
spc <- cbind.data.frame(spc, name)
colnames(spc) <- c("Species ID", "Species Name", "Common Name")
spc
```


6.  Finally, we will use a regression to look at the relationship
    between spending and species taxon.

Because the distribution of spending is very right-skewed, it would be a
good idea to take the logarithm of spending before using it in a
regression.

Remember that log(0)=infinity. That means we have to drop observations
with zero spending before taking the logarithm.

a)  Drop the rows where spending == 0 from the data frame and then make
    a new column with the logarithm (log()) of spending in each year. (2
    points)

Optional: Look at the distribution of the logged spending variable and
see how it looks different from the plot you made in question 4a

```{r}
cslog <- cosp[cosp$spending!="0",]
cslog <- transform(
  cslog, log_spending = log(cslog$spending))

ggplot(cslog, aes(x=log_spending)) +
  geom_density(fill="darkgreen", color="grey", alpha=0.8) +
  labs(title="Distribution of Spending from 2006 to 2016", x="Spending in Logarithmic Scale", y="Spending Density")

```

b)  Run a regression of logged spending on taxonomic group and print the
    summary for the regression below (3 points)

```{r}
reg <- lm(log_spending~taxon, data=cslog)
summary.lm(reg)
```

c)  The way to interpret these coefficients are as the fractional
    difference in spending between the taxonomic group (e.g. Birds,
    Fishes, etc) and the "dropped" group, where by default the dropped
    group will be Amphibians. Positive numbers indicate that group has
    more spent on it than Amphibians and negative numbers indicate it
    has less spent on it.

Based on your results in b, do we see statistically significant
differences in spending between different taxonomic groups? If so, which
kinds of species tend to have more spent on them and which have less? (1
points)

**Answer:**

<span style="color:orange">
Taxonomic group that have the most spend is **Mammals**. It is seen its Estimate number of **1.03077** that is the largest positive number of Estimate Coefficient. 
</span>

<span style="color:orange">
Taxonomic group that have the least spend is **Plants**. It is seen its Estimate number of **-1.92320** that is the largest negative number of Estimate Coefficient.
</span>

7.  Push your R markdown file to your Github repository (2 points)
